Real-Time Activity Classification by Match Filtering using Body-Worn Accelerometers

### 2016-06-20 notes/paper/paper.tex
Added more to the paper.

\documentclass[journal]{IEEEtran}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{subfig}
%\usepackage{caption}
\usepackage[english]{babel}
\usepackage{pinyin}
\usepackage{natbib}
\usepackage{notoccite}
\usepackage{amssymb}
\usepackage{textcomp}
\usepackage{floatrow}
\usepackage{amsmath}
%
\begin{document}
%
% paper title
\title{Real-Time Activity Classification by Match Filtering using Body-Worn Accelerometers}
%
\author{Craig~Euler, C.T.~Lin, Bryan~Juarez, Melissa~Flores}
%
\maketitle
%
\begin{abstract}
Body worn sensors have been used in the market for many years now.
Many of which are used for activity monitoring or fitness.
In this paper, we make use of a matched-filtering algorithm for real-time activity classification with data obtained from three-axis accelerometers worn by the user.
We achieve invariance to sensor orientation with dimensionality reduction.
We make use of an instance-based learning algorithm to train the device to learn the individual's motion patterns and store that information as activity filters for our matched filter.
We improve computational efficiency with dimensionality reduction, decimation in time, and pre-processing of our references post training for use in the matched-filter.
An assessment of our computational costs are provided with runtime experiments on the MSP432 micro-controller.
\end{abstract}
%
\section{Introduction}
Various methods for activity classification using body-worn sensors exist.
Custom decision tree, automatically generated decision tree, and artificial neural networks \cite{parkka_ermes_korpipaa_mantyjarvi_peltola_korhonen_2006} have been explored as well as other frequency based classification methods \cite{sharma_purwar_lee_lee_chung_2008}.
In this paper, we make use of a matched-filter method to identify an individual\textquotesingle s activity in real-time.
The matched-filter method with use of thresholding and comparing the correlations of various filters is a simple method to use and have been studied in the past \cite{giannakis_tsatsanis_1990}.
Many of the algorithms and software structure were tailored to computational efficiency to accomidate the resource limitations on the MSP 432 micro-processor.
Having an a priori knowledge of the activity the individual is performing is useful for many current fitness algorithms.
Devices such as the Fit-Bit, Apple watch, Moov, Jawbone, etc. usually have the individual actively input the activity they are performing.

@@

The second module is the real-time activity classification on the unit.
This module reads in the reference(s) stored on the unit to be used as templates that the matched-filter will use for determining the individual\textquotesingle s activity.
Both of these modules use the same principles and hence the same algorithm kernels.
%
\section{Algorithms}
\subsection{Activity Classification Method}
The matched-filter method is commonly used for extracting a signal out of noisy data.
We use this method for identifying a user\textquotesingle s activity from data generated by body-worn accelerometers.
Because the matched-filter method requires use of an existing reference template to identify the desired signal, we must develop our template from a training set.
We will discuss how we develop our set of templates for use in our matched-filter.

The matched filter is based on the cross correlation of the data $\textbf{y} = \{y\}_{k=0}^{N-1}$ with the reference signal $\textbf{x} = \{x\}_{k=0}^{M-1}$.
The cross correlation of $\textbf{x}$ with $\textbf{y}$ is defined as:
%
\begin{equation} \label{cross_correlation_eq}
(\textbf{x} \star \textbf{y}) = \left \{\sum_{i=0}^{M}x_{i} y_{i+k} \right \}_{k=0}^{N-M-1}
\end{equation}
%
which is simply a moving dot product.
We can then normalize \eqref{cross_correlation_eq} by the following:
%
\begin{equation} \label{norm_cross_correlation_eq}
\widehat{(\textbf{x} \star \textbf{y})}_k = \frac{(\textbf{x} \star \textbf{y})_k}{||\textbf{x}|| \ || \widetilde{\textbf{y}}_k || }, \quad k = 0,1,...,N-M-1
\end{equation}
%
where $|| \cdot ||$ is the usual $\ell_2$ norm and $\widetilde{\textbf{y}}_k = \{y_p\}_{p=k}^{M+k-1}$.
\eqref{norm_cross_correlation_eq} is now bounded by $[-1,1]$ and has any bias toward high energy removed.
We improve our computational efficiency by applying the convolution theorem to \eqref{cross_correlation_eq} and replacing it with:
%
\begin{equation} \label{conv_theorem}
(\textbf{x} \star \textbf{y}) = FFT^{-1}(FFT(\textbf{x}') FFT(\textbf{y})^*)
\end{equation}
%
Where $\textbf{x}'$ is $\textbf{x}$ that has been zero padded to the length of $\textbf{y}$ with the added restriction that the length of $\textbf{x}$ is no more than the length of $\textbf{y}$ and $FFT$ is the Fast Fourier Transform algorithm.
We are most interested where our reference $\textbf{x}$ best matches our data $\textbf{y}$.
We use the resulting max correlation to define our metric for  our matched-filter output as:
%
\begin{equation} \label{matched_filter_eq}
MF_{\textbf{x}}(\textbf{y}) := \underset{k \in [0, M-N)}{max} \left \{\widehat{(\textbf{x} \star \textbf{y})}_k \right \}
\end{equation}
%
If \eqref{matched_filter_eq} passes a threshold, the corresponding signal is a match.
%
\subsection{Dimensionality Reduction}
We manage to reduce our computational resources and remove our dependence on sensor orientation at the same time through a process of dimensionality reduction.
We assume that over a short period of time, a user\textquotesingle s limb movement is constrained mostly to a two-dimensional plane.
By identifying the plane of motion, we then project the acceleration data onto this plane and re-define our new axis accordingly.
By this assumption, activities that are mostly confined to two dimensions will improve our signal identification.

If we let $\textbf{a} = \{a_i\}_{i=0}^{N}$ represent our demeaned acceleration data over a specified time window in three dimensions then the covariance matrix of our acceleration data is represented by the real-symmetric $3 \times 3$ matrix $Q = \textbf{a} \textbf{a}^T$.
The resulting unit length eigen vectors $V = (\textbf{v}^1,\textbf{v}^2,\textbf{v}^3)$ for the respective eigen values $\lambda^1 \geq \lambda^2 \geq \lambda^3$ of this matrix are orthonormal and point in the directions where the data varies most \cite{bishop_2006}.
% \ref{bishop_2006}
\begin{equation} \label{project_eq}
   \textbf{a}' = V^T \textbf{a}
\end{equation}
%
We now represent our acceleration data as $\textbf{a}'$ after rotating our axis by \eqref{project_eq}. We make the assumption that our data is most dominate in two dimensions i.e. $\lambda^3 << \lambda^2$ so it is safe to assume that
%
\begin{equation} \label{magnitude_eq}
||(a^1,a^2,a^3)|| = ||(a'^1,a'^2,a'^3)|| \sim = (a'^1)^2 + (a'^2)^2
\end{equation}
$ \forall a \in \textbf{a}. $ \\
%
Equality comes from the orthonormality of $V$. The margin of error in the approximation is not important since we are only interested in processing in two dimensions. Our goal is only to minimized the margin of error so that we are not, by chance, using an axis direction where the data varies least.
By projecting our data onto the plain defined by the two most dominant eigen vectors, we effectively reduce our representation of our data to two dimensions.

@@

We chose this method for its simplicity and robustness against outliers in our training set. We anticipate that the user will have idle motion at the start and end of their training duration when they activate the training mode and with possible disruptions during the process.
%
\section{Performance and Throughput}
%
The Kinetisense sensor was used for the performance analysis in this paper.
This unit has a three-axis accelerometer with an acceleration range of -5g to +5g and a sampling rate of 128Hz.
Samples of walking, jogging, and bicycling have been recorded with this unit attached to the ankle of the user for the purpose of analyzing the performance.
A separate instance of walking data has been used specifically for training the classifier.
In figure \ref{fig:MF_performance}, \subref{fig:frac_walking_jogging} and \subref{fig:frac_walking_bicycling} show the fraction of the correlation results that exceed the given correlation threshold when comparing the walking data reference to the given activity.
As we would expect, walking data matches better to walking data than the other activities.
ROC curves were produced from these corresponding results in \subref{fig:roc_walking_jogging} and \subref{fig:roc_walking_bicycling} to illistrate the effectiveness of the matched-filter as a classifier.
Figure \ref{fig:MF_performance} shows the comparison to bicycling.
Because the bicycling motions are considerably different from walking relative to jogging, we would expect better performance from bicycling.
%
%\begin{figure}[!tbp]
%\RawFloats
%  \centering
%     \begin{minipage}[b]{0.45\textwidth}
%        \includegraphics[angle=0,width=\textwidth]{ROC_walk_to_run.png}
%        \caption{\label{ROC_walk_to_run_fig}ROC curve with a downsample factor of 8}
%     \end{minipage}
%     \hfill
%     \begin{minipage}[b]{0.45\textwidth}
%     \includegraphics[angle=0,width=\textwidth]{ROC_walk_to_bike.png}
%        \caption{\label{ROC_walk_to_bike_fig}ROC curve with a downsample factor of 8}
%     \end{minipage}
%\caption{\label{ROC_walk_to_bike_fig}ROC curve with a downsample factor of 8}
%\end{figure}
%
\begin{figure}[!ht]
  %\captionsetup[subfigure]{labelformat=empty}
  \centering
  \subfloat[\label{fig:frac_walking_jogging}]{\includegraphics[width=.45\textwidth]{fraction_threshold_walking_to_jogging.png}}
  \quad
  \subfloat[\label{fig:frac_walking_bicycling}]{\includegraphics[width=.45\textwidth]{fraction_threshold_walking_to_bicycling.png}}
  \centering
  \quad
  \subfloat[\label{fig:roc_walking_jogging}]{\includegraphics[width=.45\textwidth]{ROC_walk_to_run.png}}
  \centering
  \quad
  \subfloat[\label{fig:roc_walking_bicycling}]{\includegraphics[width=.45\textwidth]{ROC_walk_to_bike.png}}
  \caption{Matched-filter performance comparing jogging and bicycling data against a reference for walking. Results produced with decimation 8}
  \label{fig:MF_performance}
\end{figure}
%
%\FloatBarrier
reference to figure \ref{fig:MF_performance}
%
\begin{figure}[!ht]
   \centering
   \includegraphics[width=1.0\textwidth]{one_and_two_references.png} \label{fig:timing_refs}
   \caption{Timing results with varying sampling frequencies on the MSP432 micro-controller processor comparing the timing between using one vs two references for classification purposes. The time window used for activity classification is 4.0 seconds}
\end{figure}
%
\section{Conclusion}
conclusion text.
%
\appendices
%
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi
%
\bibliographystyle{plain}
\bibliography{reference}
\end{document}

>>>>

\documentclass[journal]{IEEEtran}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{subfig}
%\usepackage{caption}
\usepackage[english]{babel}
\usepackage{pinyin}
\usepackage{natbib}
\usepackage{notoccite}
\usepackage{amssymb}
\usepackage{textcomp}
\usepackage{floatrow}
\usepackage{amsmath}
%
\begin{document}
%
% paper title
\title{Real-Time Activity Classification by Match Filtering using Body-Worn Accelerometers}
%
\author{Craig~Euler, C.T.~Lin, Bryan~Juarez, Melissa~Flores}
%
\maketitle
%
\begin{abstract}
Body worn sensors have been used in the market for many years now.
Many of which are used for activity monitoring or fitness.
In this paper, we make use of a matched-filtering algorithm for real-time activity classification with data obtained from three-axis accelerometers worn by the user.
We achieve invariance to sensor orientation with dimensionality reduction.
We make use of an instance-based learning algorithm to train the device to learn the individual's motion patterns and store that information as activity filters for our matched filter.
We improve computational efficiency with dimensionality reduction, decimation in time, and pre-processing of our references post training for use in the matched-filter.
An assessment of our processing time is provided with runtime experiments on the MSP432 micro-controller which shows feasability of real-time processing capability.
\end{abstract}
%
\section{Introduction}
Various methods for activity classification using body-worn sensors exist.
Custom decision tree, automatically generated decision tree, and artificial neural networks \cite{parkka_ermes_korpipaa_mantyjarvi_peltola_korhonen_2006} have been explored as well as other frequency based classification methods \cite{sharma_purwar_lee_lee_chung_2008}.
In this paper, we make use of a matched-filter method to identify an individual\textquotesingle s activity in real-time.
The matched-filter method with use of thresholding and comparing the correlations of various filters is a simple method to use and have been studied in the past \cite{giannakis_tsatsanis_1990}.
Many of the algorithms and software structure were tailored to computational efficiency to accomidate the resource limitations on the MSP 432 micro-processor.
Having an a priori knowledge of the activity the individual is performing is useful for many current fitness algorithms.
Devices such as the Fit-Bit, Apple watch, Moov, Jawbone, etc. usually have the individual actively input the activity they are performing.

@@

The second module is the real-time activity classification on the unit.
This module reads in the reference(s) stored on the unit to be used as templates that the matched-filter will use for determining the individual\textquotesingle s activity.
Both of these modules use the same principles and hence the same algorithm kernels.
%
\section{The Algorithms}
\subsection{Activity Classification Method}
The matched-filter method is commonly used for extracting a signal out of noisy data.
We use this method for identifying a user\textquotesingle s activity from data generated by body-worn accelerometers.
Because the matched-filter method requires use of an existing reference template to identify the desired signal, we must develop our template from a training set.
We will discuss how we develop our set of templates for use in our matched-filter.

The matched filter is based on the cross correlation of the data $\textbf{y} = \{y\}_{k=0}^{N-1}$ with the reference signal $\textbf{x} = \{x\}_{k=0}^{M-1}$.
The cross correlation of $\textbf{x}$ with $\textbf{y}$ is defined as:
%
\begin{equation} \label{cross_correlation_eq}
(\textbf{x} \star \textbf{y}) = \left \{\sum_{i=0}^{M}x_{i} y_{i+k} \right \}_{k=0}^{N-M-1}
\end{equation}
%
which is simply a moving dot product.
We can then normalize \eqref{cross_correlation_eq} by the following:
%
\begin{equation} \label{norm_cross_correlation_eq}
\widehat{(\textbf{x} \star \textbf{y})}_k = \frac{(\textbf{x} \star \textbf{y})_k}{||\textbf{x}|| \ || \widetilde{\textbf{y}}_k || }, \quad k = 0,1,...,N-M-1
\end{equation}
%
where $|| \cdot ||$ is the usual $\ell_2$ norm and $\widetilde{\textbf{y}}_k = \{y_p\}_{p=k}^{M+k-1}$.
\eqref{norm_cross_correlation_eq} is now bounded by $[-1,1]$ and has any bias toward high energy removed.
We improve our computational efficiency by applying the convolution theorem to \eqref{cross_correlation_eq} and replacing it with:
%
\begin{equation} \label{conv_theorem}
(\textbf{x} \star \textbf{y}) = FFT^{-1}(FFT(\textbf{x}') FFT(\textbf{y})^*)
\end{equation}
%
Where $\textbf{x}'$ is $\textbf{x}$ that has been zero padded to the length of $\textbf{y}$ with the added restriction that the length of $\textbf{x}$ is no more than the length of $\textbf{y}$ and $FFT$ is the Fast Fourier Transform algorithm.
We are most interested where our reference $\textbf{x}$ best matches our data $\textbf{y}$.
We use the resulting max correlation to define our metric for  our matched-filter output as:
%
\begin{equation} \label{matched_filter_eq}
MF_{\textbf{x}}(\textbf{y}) := \underset{k \in [0, M-N)}{max} \left \{\widehat{(\textbf{x} \star \textbf{y})}_k \right \}
\end{equation}
%
If \eqref{matched_filter_eq} passes a threshold, the corresponding signal is a match.
%
\subsection{Dimensionality Reduction}
We manage to reduce our computational resources and remove our dependence on sensor orientation at the same time through a process of dimensionality reduction with principal component analaysis (PCA) \cite{bishop_2006}.
We assume that over a short period of time, a user\textquotesingle s limb movement is constrained mostly to a two-dimensional plane.
By identifying this plane of motion, we then project the acceleration data onto this plane and re-define our new axis accordingly.

If we let $\textbf{a} = \{a_i\}_{i=0}^{N}$ represent our demeaned acceleration data over a specified time window in three dimensions then the covariance matrix of our acceleration data is represented by the real-symmetric $3 \times 3$ matrix $Q = \textbf{a} \textbf{a}^T$.
The resulting unit length eigen vectors $V = (\textbf{v}^1,\textbf{v}^2,\textbf{v}^3)$ with their respective eigen values $\lambda^1 \geq \lambda^2 \geq \lambda^3$ of this matrix are orthonormal and point in the directions where the data varies most in descending order for each orthogonal direction.
% \ref{bishop_2006}
\begin{equation} \label{project_eq}
   \textbf{a}' = V^T \textbf{a}
\end{equation}
%
We now represent our acceleration data as $\textbf{a}'$ after projecting our acceleration data $\textbf{a}$ onto our rotated axis $V$ by \eqref{project_eq}. We make the assumption that our data is most dominate in two dimensions i.e. $\lambda^3 << \lambda^2$ so it is safe to assume that
%
\begin{equation} \label{magnitude_eq}
||(a^1,a^2,a^3)|| = ||(a'^1,a'^2,a'^3)|| \sim = (a'^1)^2 + (a'^2)^2
\end{equation}
$ \forall a \in \textbf{a}. $ \\
%
Equality comes from the orthonormality of $V$. The margin of error in the approximation is not important since we are only interested in processing in the two dimensions.
By now we have preserved as much information that is contained in our acceleration data as possible while also reducing our dimension from three to two.
By projecting our data onto the plain defined by the two most dominant eigen vectors, we effectively reduce our representation of our data to two dimensions.

@@

We chose this method for its simplicity and robustness against outliers in our training set. We anticipate that the user will have idle motion at the start and end of their training duration when they activate the training mode and with possible disruptions during the process.
%
\section{Performance and Throughput}
%
The Kinetisense sensor was used for the performance analysis in this paper.
This unit has a three-axis accelerometer with an acceleration range of -5g to +5g and a sampling rate of 128Hz.
Samples of walking, jogging, and bicycling have been recorded with this unit attached to the ankle of the user for the purpose of analyzing the performance.
A separate instance of walking data has been used specifically for training the classifier.
In figure \ref{fig:MF_performance}, \subref{fig:frac_walking_jogging} and \subref{fig:frac_walking_bicycling} show the fraction of the correlation results that exceed the given correlation threshold when comparing the walking data reference to the given activity.
As we would expect, walking data matches better to walking data than the other activities.
ROC curves were produced from these corresponding results in \subref{fig:roc_walking_jogging} and \subref{fig:roc_walking_bicycling} to illistrate the effectiveness of the matched-filter as a classifier.
Figure \ref{fig:MF_performance} shows the comparison to bicycling.
Because the bicycling motions are considerably different from walking relative to jogging, we would expect better performance from bicycling.
%
%\begin{figure}[!tbp]
%\RawFloats
%  \centering
%     \begin{minipage}[b]{0.45\textwidth}
%        \includegraphics[angle=0,width=\textwidth]{ROC_walk_to_run.png}
%        \caption{\label{ROC_walk_to_run_fig}ROC curve with a downsample factor of 8}
%     \end{minipage}
%     \hfill
%     \begin{minipage}[b]{0.45\textwidth}
%     \includegraphics[angle=0,width=\textwidth]{ROC_walk_to_bike.png}
%        \caption{\label{ROC_walk_to_bike_fig}ROC curve with a downsample factor of 8}
%     \end{minipage}
%\caption{\label{ROC_walk_to_bike_fig}ROC curve with a downsample factor of 8}
%\end{figure}
%
\begin{figure}[!ht]
  %\captionsetup[subfigure]{labelformat=empty}
  \centering
  \subfloat[\label{fig:frac_walking_jogging}]{\includegraphics[width=.45\textwidth]{fraction_threshold_walking_to_jogging.png}}
  \quad
  \subfloat[\label{fig:frac_walking_bicycling}]{\includegraphics[width=.45\textwidth]{fraction_threshold_walking_to_bicycling.png}}
  \centering
  \quad
  \subfloat[\label{fig:roc_walking_jogging}]{\includegraphics[width=.45\textwidth]{ROC_walk_to_run.png}}
  \centering
  \quad
  \subfloat[\label{fig:roc_walking_bicycling}]{\includegraphics[width=.45\textwidth]{ROC_walk_to_bike.png}}
  \caption{Matched-filter performance comparing jogging and bicycling data against a reference for walking. Results produced with decimation 8}
  \label{fig:MF_performance}
\end{figure}
%
%\FloatBarrier
reference to figure \ref{fig:MF_performance}
%
\begin{figure}[!ht]
   \centering
   \includegraphics[width=1.0\textwidth]{one_and_two_references.png} \label{fig:timing_refs}
   \caption{Timing results with varying sampling frequencies on the MSP432 micro-controller processor comparing the timing between using one vs two references for classification with the matched-filter. The time window used for activity classification is 4.0 seconds}
\end{figure}
%
\section{Conclusion}
conclusion text.
%
\appendices
%
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi
%
\bibliographystyle{plain}
\bibliography{reference}
\end{document}
